A Speculative Approach to Parallelization in Particle Swarm Optimization

### 2011-04-01 masters_thesis/si_paper/paper.tex
A few minor changes.  Basically done.

The balance of this paper is organized as follows. \sec{pso} describes the
particle swarm optimization algorithm, and \sec{related} gives a brief overview
of previous parallelization techniques for this algorithm.  \sec{sepso} shows
mathematically how speculative evaluation can be done in parallel PSO to
perform two iterations at once, then \sec{implementation} gives information on
how to implement this method.  In \sec{relax}, we discuss various methods of
improving the performance of speculative evaluation in PSO, all of which break
the requirement of strictly reproducing the behavior of the original algorithm.
\sec{setup} describes the experiments we ran, and \sec{results} presents the
results of those experiments.  In \sec{conclusion} and \sec{future} we conclude
and discuss future work.

@@

Becuase there are a few branches with very high probabilities in the topologies
we are interested in, we can have hope that cutting out some of the evaluations
that have low probability will lead to an increase in overall performance.  In
order to implement this kind of pruning, the only change that needs to be made
to \alg{centralized} is in Step 2, where the speculative children are
generated.  Instead of generating all possible speculative children, generate
the subset of the children that is desired.

@@

In parallel PSO on long-running functions, the natural way to present results
is in terms of iterations, not function evaluations.  This is because when all
function evaluations at each iteration are performed concurrently, iterations
are the direct equivalent of wall-clock time.  Thus we report iterations in our
results instead of function evaluations.  But because SEPSO actually performs
two iterations of PSO at each ``iteration,'' and Social Promotion and Many
Iterations make the idea of ``iterations of PSO'' somewhat nebulus, we instead
call each ``iteration'' a ``time step.''

@@

We ran 50 trials of each experiment with Griewank, so that the curves are more
smooth.  We show results in \fig{basic-griewank1} for swarms of size 100 and
800 using the Ring topology.  One can see in the figure that SEPSO reaches the
global optimum on average close to twice as fast as Standard, while Pick Best
is close to three times as fast and Many Iterations is even faster when it is
successful, though it is only successful 84\% of the time.  \tabref{griewank}
shows all of our results in tabular form.

@@

@@

\subsection{Terminology}

@@

\subsection{Centralized Algorithms}

@@

\subsection{Distributed Algorithms}

@@

\subsection{Dynamic Topologies}

>>>>

The balance of this paper is organized as follows. \sec{pso} describes the
particle swarm optimization algorithm, and \sec{related} gives a brief overview
of previous parallelization techniques for this algorithm.  \sec{sepso} shows
mathematically how speculative evaluation can be done in parallel PSO to
perform two iterations at once, leaving implementation concerns to the
appendices.  In \sec{relax}, we discuss various methods of improving the
performance of speculative evaluation in PSO, all of which break the
requirement of strictly reproducing the behavior of the original algorithm.
\sec{setup} describes the experiments we ran, and \sec{results} presents the
results of those experiments.  In \sec{conclusion} and \sec{future} we conclude
and discuss future work.

@@

Because there are a few branches with very high probabilities in the topologies
we are interested in, we can have hope that cutting out some of the evaluations
that have low probability will lead to an increase in overall performance.  In
order to implement this kind of pruning, the only change that needs to be made
to \alg{centralized} is in Step 2, where the speculative children are
generated.  Instead of generating all possible speculative children, generate
the subset of the children that is desired.

@@

In parallel PSO on long-running functions, the natural way to present results
is in terms of iterations, not function evaluations.  This is because when all
function evaluations at each iteration are performed concurrently, iterations
are the direct equivalent of wall-clock time.  Thus we report iterations in our
results instead of function evaluations.  But because SEPSO actually performs
two iterations of PSO at each ``iteration,'' and Social Promotion and Many
Iterations make the idea of ``iterations of PSO'' somewhat nebulous, we instead
call each ``iteration'' a ``time step.''

@@

We ran 50 trials of each experiment with Griewank, so that the curves are more
smooth.  We show results in \fig{basic-griewank1} for swarms of size 100 and
800 using the Ring topology.  One can see in the figure that SEPSO reaches the
global optimum on average close to twice as fast as Standard, while Pick Best
is close to three times as fast and Many Iterations is even faster when it is
successful, though it is only successful 84\% of the time.
\tabref{griewank-20} shows all of our results in tabular form.

@@

Note here that \fig{griewank-manyiters} is interesting to compare to
\fig{basic-griewank1}.  The ordering of the methods in terms of the number of
time steps to completion is the same in both figures.  What is different is
that at 240 processors, most of our methods fail to find the optimum 100\% of
the time, while at 800 processors, all but Many Iterations succeed 100\% of the
time, and Many Iterations is very close.  At some point between 240 and 800
processors, SEPSO and Pick Best become successful, and at that point it is by
far better to use speculative methods than standard parallelizations.  We can
also assume that at some point above 800 processors, Many Iterations will be
become successful and will be the preferred method.  Once there are enough
particles in the swarm to guarantee success when optimizing a function, the
best use of additional processors is to speculate as far ahead as possible,
speeding up the progress of the algorithm.

@@

\subsection*{Terminology}

@@

\subsection*{Centralized Algorithms}

@@

\subsection*{Distributed Algorithms}

@@

\subsection*{Dynamic Topologies}
